---
title: "cns_early_voting_taylor"
author: "Taylor Nichols"
date: "2024-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(tidycensus)
library(scales)
```

```{r}
early_voting <- read_csv("raw_data_nov_1.csv") |>
  clean_names()

early_voting_2016 <- read_csv("PG16 EarlyVoting RAW data.csv")|>
  clean_names()

early_voting <- early_voting |>
  rowwise() %>%
  mutate(total_votes = sum(c_across(day1:day8), na.rm = TRUE)) %>%
  ungroup()

early_voting_2016 <- early_voting_2016 |>
  rowwise() %>%
  mutate(total_votes = sum(c_across(day1:day8), na.rm = TRUE)) %>%
  ungroup()

early_voting_grouped <- early_voting |>
  group_by(county_name, gender_code, party_code, age_group) %>%
  summarize(total_votes_sum = sum(total_votes, na.rm = TRUE)) %>%
  ungroup()

early_voting_grouped_2016 <- early_voting_2016 |>
  group_by(county_name, gender_code, party_code, age_group) %>%
  summarize(total_votes_sum = sum(total_votes, na.rm = TRUE)) %>%
  ungroup()
```


```{r}
early_voting_grouped |>
  group_by(county_name) |>
  summarise(total_votes_2024 = sum(total_votes_sum)) |> 
  arrange(desc(total_votes_2024))

early_voting_grouped_2016 |>
  group_by(county_name) |>
  summarise(total_votes = sum(total_votes_sum)) |> 
  arrange(desc(total_votes))
```


```{r}
# Get total population and population 18 and older
adult_pop <- get_acs(geography =  "county",
                     state = "24",
                     variables = c(adults_18_plus = "B01001_026"),
                     year = 2022)

adult_pop <- adult_pop |>
    mutate(NAME = case_when(
        NAME == "Baltimore County, Maryland" ~ str_replace(NAME, ", Maryland", ""),  # Trim ", Maryland" for Baltimore
        TRUE ~ str_replace(NAME, "County, Maryland", "")  # Trim "County, Maryland" for the rest
    )) |>
    mutate(NAME = str_replace(NAME, "city, Maryland", "City"),  # Correct case sensitivity
           NAME = str_replace(NAME, "St.", "Saint")) |>
    select(NAME, estimate) |>
    rename(county_name = NAME, population = estimate) |>
    mutate(county_name = str_trim(county_name))
```

Questions to answer: 
Who is voting? Gender/age? Gender/party? Gender/county? 
gender/age/party 
gender/age/county 
gender/party/county 
gender/age/party/county
age/party? age/county? 
age/party/county
party/county?

1. Gender/age? Gender/party?
```{r}
# Summarising gender and age groups
gender_age <- early_voting_grouped |>
  group_by(gender_code, age_group) |>
  summarise(total_votes = sum(total_votes_sum, na.rm = TRUE))

gender_age_2016 <- early_voting_grouped_2016 |>
  group_by(gender_code, age_group) |>
  summarise(total_votes_2016 = sum(total_votes_sum, na.rm = TRUE))

# Summarising gender and party groups
gender_party <- early_voting_grouped |>
  filter(party_code %in% c("DEM", "REP", "UNA")) |>
  group_by(gender_code, party_code) |>
  summarise(total_votes = sum(total_votes_sum, na.rm = TRUE))

gender_party_2016 <- early_voting_grouped_2016 |>
  filter(party_code %in% c("DEM", "REP", "UNA")) |>
  group_by(gender_code, party_code) |>
  summarise(total_votes_2016 = sum(total_votes_sum, na.rm = TRUE))

# Summarising gender, age, and party groups
gender_age_party <- early_voting_grouped |>
  filter(party_code %in% c("DEM", "REP", "UNA")) |>
  group_by(gender_code, age_group, party_code) |>
  summarise(total_votes = sum(total_votes_sum, na.rm = TRUE))

# Summarising age and party groups
age_party <- early_voting_grouped |>
  filter(party_code %in% c("DEM", "REP", "UNA")) |>
  group_by(age_group, party_code) |>
  summarise(total_votes = sum(total_votes_sum, na.rm = TRUE))

age_party_2016 <- early_voting_grouped_2016 |>
  filter(party_code %in% c("DEM", "REP", "UNA")) |>
  group_by(age_group, party_code) |>
  summarise(total_votes_2016 = sum(total_votes_sum, na.rm = TRUE))

# Merging data frames
gender_age <- full_join(gender_age, gender_age_2016, by = c("gender_code", "age_group"))
age_party <- full_join(age_party, age_party_2016, by = c("party_code", "age_group"))
gender_party <- full_join(gender_party, gender_party_2016, by = c("gender_code", "party_code"))

# Calculate the total sum of total_votes for all voters
all_votes_2024 <- sum(early_voting_grouped$total_votes_sum, na.rm = TRUE)
all_votes_2016 <- sum(early_voting_grouped_2016$total_votes_sum, na.rm = TRUE)

gender_party <- gender_party |>
  mutate(
    pct = total_votes / all_votes_2024 * 100,
    pct_2016 = total_votes_2016 / all_votes_2016 * 100
  )

age_party <- age_party |>
  mutate(
    pct = total_votes / all_votes_2024 * 100,
    pct_2016 = total_votes_2016 / all_votes_2016 * 100
  )
# Assuming all_votes_2024 and all_votes_2016 are defined somewhere above this code
gender_age <- gender_age |>
  mutate(
    pct = total_votes / all_votes_2024 * 100,
    pct_2016 = total_votes_2016 / all_votes_2016 * 100
  )

# Optionally, round the percentages for better readability
gender_age <- gender_age |>
  mutate(
    pct = round(pct, 2),       # Round to 2 decimal places
    pct_2016 = round(pct_2016, 2)  # Round to 2 decimal places
  )

```


```{r}
view(gender_age)
view(gender_party)
view(age_party)
```


```{r}
early_voting_grouped <- early_voting_grouped |>
rename(total_votes_2024 = total_votes_sum)

early_voting_grouped_2016 <- early_voting_grouped_2016 |>
rename(total_votes_2016 = total_votes_sum)

early_voting_grouped_full <- left_join(early_voting_grouped, early_voting_grouped_2016, 
                                       by = c("county_name", "gender_code", "party_code", "age_group"))
```


```{r}
early_voting_grouped_full |>
  group_by(gender_code) |>
  summarise(
    total_votes_2024 = sum(total_votes_2024, na.rm = TRUE),
    total_votes_2016 = sum(total_votes_2016, na.rm = TRUE)) |>
  mutate(pct_2024 = total_votes_2024/all_votes_2024, pct_2016 = total_votes_2016/all_votes_2016)

early_voting_grouped_full |>
  group_by(age_group) |>
  summarise(
    total_votes_2024 = sum(total_votes_2024, na.rm = TRUE),
    total_votes_2016 = sum(total_votes_2016, na.rm = TRUE)) |>
  mutate(pct_2024 = total_votes_2024/all_votes_2024, pct_2016 = total_votes_2016/all_votes_2016)

```

```{r}
# Reshape data to have a single 'total_votes' column with a 'year' identifier
gender_age_long <- early_voting_grouped_full |>
  group_by(age_group, gender_code) |>
  summarise(
    total_votes_2024 = sum(total_votes_2024, na.rm = TRUE),
    total_votes_2016 = sum(total_votes_2016, na.rm = TRUE)
  ) |>
  pivot_longer(cols = c(total_votes_2024, total_votes_2016),
               names_to = "year",
               values_to = "total_votes") |>
  mutate(year = ifelse(year == "total_votes_2024", "2024", "2016"))

# Plotting
ggplot(gender_age_long, aes(x = age_group, y = total_votes, fill = gender_code)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ year) +
  labs(title = "Counts by Age Group, Gender, and Election Year",
       x = "Age Group",
       y = "Total Votes") +
  theme_minimal() +
  scale_y_continuous(labels = comma)

```

```{r}
ggplot(gender_age, aes(x = age_group, y = total_votes, fill = gender_code)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Counts by Age Group and Gender",
       x = "Age Group",
       y = "Count") +
  theme_minimal() +
  scale_y_continuous(labels = comma)

ggplot(gender_party, aes(x = party_code, y = total_votes, fill = gender_code)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Counts by Party and Gender",
       x = "Party Code",
       y = "Count") +
  theme_minimal() +
  scale_y_continuous(labels = comma)

ggplot(gender_age_party, aes(x = interaction(gender_code, age_group, party_code), y = total_votes, fill = party_code)) +
  geom_bar(stat = "identity") +
  labs(title = "Counts by Gender, Age, and Party",
       x = "Gender-Age-Party Grouping",
       y = "Count") +
  theme_minimal() +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates x-axis labels for readability

ggplot(age_party, aes(x = party_code, y = total_votes, fill = age_group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Counts by Party and Age",
       x = "Party Code",
       y = "Count") +
  theme_minimal() +
  scale_y_continuous(labels = comma)

```
Overall findings: 
45-64 year olds are turning out in higher numbers, followed by folks 65 and up. Women in both these age ranges are turning out in higher rates than men, with women ages 45-64 leading followed by women ages 65 and up. Democrat women are the largest group by a wide margin (more than 7,000 voters), followed by democrat men.

For Republicans, men and women ages 45+ are turning out at higher rates. Younger Republicans aren't turning out early at very high rates at all. 

gender/age/county 
gender/party/county 
gender/age/party/county
age/party? age/county? 
age/party/county
party/county?

```{r}
age_county <- early_voting_grouped  |>
  group_by(age_group, county_name) |>
 summarise(total_votes = sum(total_votes_2024)) |> 
  arrange(desc(total_votes))

gender_party_county <- early_voting_grouped  |>
  filter(party_code %in% c("DEM", "REP", "UNA")) |>
  group_by(gender_code, party_code, county_name) |>
 summarise(total_votes = sum(total_votes_2024)) |> 
  arrange(desc(total_votes))

age_county <- left_join(age_county, adult_pop, by = "county_name") |>
  mutate(per_10k = total_votes/population*10000)

gender_party_county <- left_join(gender_party_county, adult_pop, by = "county_name") |>
  mutate(per_10k = total_votes/population*10000)

age_county |>
  arrange(desc(per_10k))

gender_party_county |>
  arrange(desc(per_10k))
```

Overall, it looks like older adults in Talbot and Queen Anne's counties are having the highest turnout per capita, with between 1300-1600 voters per 10k or about 130-160 voters per 1k of the adult population. After that, voters 45-64 in Calvert and Queen Anne's counties are turning out at high rates per capita.

Republican women and men in Queen Anne's are actually the highest turnout per capita, with 99-100 per 1000 people from these age groups voting early (roughly 10% of the population?). Women democrats in Charles and PG county are also having high turnout per capita, followed by republicans in Garrett, Calvert and Talbot counties.


### Early voting centers analysis
```{r}
early_voting_grouped_full_counties <- early_voting_grouped_full |>
  group_by(county_name) |>
  summarise(
    total_votes_2024 = sum(total_votes_2024, na.rm = TRUE),
    total_votes_2016 = sum(total_votes_2016, na.rm = TRUE))

early_voting_grouped_full_counties <- left_join(early_voting_grouped_full_counties, adult_pop, by = "county_name")

early_voting_grouped_full_counties <- early_voting_grouped_full_counties |>
  mutate(pct2024 = total_votes_2024/population, pct2016 = total_votes_2016/population, difference = pct2024-pct2016) |>
  arrange(county_name)
```

```{r}
age_comparison <- early_voting_grouped_full |>
  group_by(age_group) |>
  summarise(
    total_votes_2024 = sum(total_votes_2024, na.rm = TRUE),
    total_votes_2016 = sum(total_votes_2016, na.rm = TRUE))

# Reshape data for plotting
age_comparison_long <- age_comparison |>
  pivot_longer(cols = c(total_votes_2024, total_votes_2016),
               names_to = "year",
               values_to = "total_votes") |>
  mutate(year = ifelse(year == "total_votes_2024", "2024", "2016"))

# Plotting
ggplot(age_comparison_long, aes(x = age_group, y = total_votes, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Total Votes by Age Group for 2024 and 2016",
       x = "Age Group",
       y = "Total Votes") +
  theme_minimal() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("2024" = "blue", "2016" = "red"))
```


```{r}
write_csv(early_voting_grouped_full_counties, "county_changes.csv")
```

```{r}
(all_votes_2024 - all_votes_2016)/all_votes_2016*100

```

```{r}
party <- early_voting_grouped_full |>
  group_by(party_code) |>
  summarise(
party_votes_2024 = sum(total_votes_2024, na.rm = TRUE),
party_votes_2016 = sum(total_votes_2016, na.rm = TRUE)) |>
  mutate(pct_2024 = party_votes_2024/all_votes_2024, pct_2016 = party_votes_2016/all_votes_2016)

```

```{r}
write_csv(party, "party.csv")
```

```{r}
all_votes_2016
all_votes_2024
```

