---
title: "early_voting_story"
author: "daranee balachandar & taylor nichols"
date: "2024-10-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}
# Turn off scientific notation
# load your packages
options(scipen=999)
library(readr)
library(tidyr)
library(tidyverse)
library(lubridate)
library(refinr)
library(janitor)
library(rvest)
library(httr)
library(dplyr)
```

## Load early votes data
This data is as of Day 5, Sunday, Oct. 28. Early voting in Maryland is between Oct. 24 to Oct. 31.
```{r}
earlyvoting_raw_data <- read_csv("EarlyVoting RAW data.csv")
View(earlyvoting_raw_data)
```

## Clean data
```{r}
# cleaning function
earlyvoting_raw_data |>
  clean_names()
```

**Task 1: Show the votes for each county for each day**
```{r}
# Convert from wide to long format for easier aggregation
long_data <- earlyvoting_raw_data |> 
  pivot_longer(cols = starts_with("day"), 
               names_to = "day", 
               values_to = "votes")

# Calculate the total votes per county for each day
votes_per_county_per_day <- long_data |> 
  group_by(COUNTY_NAME, day) |> 
  summarize(total_votes = sum(votes, na.rm = TRUE), .groups = 'drop')

# View the results
print(votes_per_county_per_day)
```


**Task 2: Show which county has the highest early votes**
```{r}

```



**Task 3: Which day has the highest early vote turnouts**


**Task 4: What is the breakdown of women voters and men voters**

**Task 5: What is the breakdown of DEM and REP and UNA voters**


