---
title: "early_voting_story"
author: "daranee balachandar & taylor nichols"
date: "2024-10-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}
# Turn off scientific notation
# load your packages
options(scipen=999)
library(readr)
library(stringr)
library(tidyr)
library(tidyverse)
library(tidycensus)
library(lubridate)
library(refinr)
library(janitor)
library(rvest)
library(httr)
library(dplyr)
library(ggplot2)
```

## Load early votes data

This data is as of Day 5, Sunday, Oct. 28. Early voting in Maryland is between Oct. 24 to Oct. 31.

```{r}
earlyvoting_raw_data <- read_csv("raw_data_nov_1.csv")
View(earlyvoting_raw_data)

```

## Clean data

```{r}
# cleaning function
earlyvoting_raw_data <- earlyvoting_raw_data |>
  clean_names()
```

**Task 1: Show the votes for each county for each day**

```{r}
# Convert from wide to long format for easier aggregation
long_data <- earlyvoting_raw_data |> 
   pivot_longer(cols = starts_with("day"), 
              names_to = "day", 
              values_to = "votes")

# Calculate the total votes per county for each day
votes_per_county_per_day <- long_data |> 
   group_by(county_name, day) |> 
   summarize(total_votes = sum(votes, na.rm = TRUE), .groups = 'drop')

# View the results
print(votes_per_county_per_day)
```

**Task 2: Breakdown of votes according to county. upload census and compare by 10,000 per capita**

```{r}
# Load census
census_api_key("6c6f00214ef0eb2cab5b4ff78cf0bd45c520fec3", install=TRUE, overwrite=TRUE)
```

**Task 3: Show which county has the highest early votes in the first 7 days**

```{r}
county_totals<- votes_per_county_per_day |> 
  group_by(county_name) |> 
  summarise(total_early_votes = sum(total_votes)) |> 
  arrange(desc(total_early_votes))
  
county_totals
```

```{r}
census_data_2024 <- get_acs(
  geography = "county",
  variables ="B01001_026",
  year = 2022,
  state = "MD",
  survey = "acs5"
)

census_data_2024 <- census_data_2024 |> 
  rename(county_name = NAME, population = estimate)
  

print(census_data_2024)
```

```{r}
census_data_2024 <- census_data_2024 |>
  mutate(county_name = if_else(
    county_name %in% c("Baltimore City", "Baltimore County"),
    county_name,
    str_replace(county_name, " County, Maryland", "") 
  ))

census_data_2024
```

```{r}
census_data_2024 <- census_data_2024 |> 
  mutate(county_name = case_when(
    county_name == "Baltimore" ~ "Baltimore County",
    county_name == "Baltimore city, Maryland" ~ "Baltimore City",
    county_name == "St. Mary's" ~ "Saint Mary's", 
    TRUE ~ county_name  
  ))

census_data_2024
```

```{r}
county_pop_vote_early <- county_totals |> 
  left_join(census_data_2024, by = "county_name")

county_pop_vote_early
```

```{r}
county_vote_breakdown_pop <- county_pop_vote_early |> 
  mutate(votes_per_10000 = (total_early_votes / population) * 10000) |> 
  arrange(desc(votes_per_10000))

county_vote_breakdown_pop

```
```{r}
write.csv(county_vote_breakdown_pop, "county_vote_breakdown_pop.csv", row.names = FALSE)
```


```{r}
ggplot(county_vote_breakdown_pop, aes(x = reorder(county_name, votes_per_10000), y = votes_per_10000)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Early Votes per 10,000 People by County",
    x = "County",
    y = "Votes per 10,000 People"
  ) +
  theme_minimal()
```

**early votes by party and county**

```{r}
# Check the structure and column names of the dataset
str(earlyvoting_raw_data)
colnames(earlyvoting_raw_data)
```
ˆ
```{r}
census_data_2024
```

```{r}
total_votes_by_county_party <- earlyvoting_raw_data |> 
  filter(party_code %in% c("DEM", "REP", "UNA")) |> 
  group_by(county_name, party_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes)) 

total_votes_per_capita <- total_votes_by_county_party |> 
  left_join(census_data_2024, by = "county_name") |> 
  mutate(votes_per_10000 = (total_votes / population) * 10000) |> 
  select(county_name, party_code, total_votes, votes_per_10000)

total_votes_per_capita 

```


```{r}
# Group by county and party and calculate total votes per 10,000
total_votes_by_county_party <- earlyvoting_raw_data |> 
  filter(party_code %in% c("DEM", "REP", "UNA")) |> 
  group_by(county_name, party_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes)) 

# Calculate votes per 10,000 and reshape the data
total_votes_per_capita <- total_votes_by_county_party |> 
  left_join(census_data_2024, by = "county_name") |> 
  mutate(votes_per_10000 = (total_votes / population) * 10000) |> 
  select(county_name, party_code, votes_per_10000) |> 
  pivot_wider(names_from = party_code, values_from = votes_per_10000, names_prefix = "votes_per_10000_") |> 
  mutate(total_votes_per_10000 = rowSums(across(starts_with("votes_per_10000_")), na.rm = TRUE)) |> 
  arrange(desc(total_votes_per_10000))


# View the reshaped data
total_votes_per_capita

write.csv(total_votes_per_capita, "total_votes_per_capita.csv", row.names = FALSE)

```



```{r}
ggplot(total_votes_per_capita, aes(x = county_name, y = votes_per_10000, fill = party_code)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Early Votes per 10,000 Residents by County and Party",
       x = "County",
       y = "Votes per 10,000 Residents") +
  scale_fill_manual(values = c("DEM" = "blue", "REP" = "red", "UNA" = "darkgreen")) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```


**What is the breakdown of women voters and men voters**

```{r}
vote_early_data <- earlyvoting_raw_data |> 
  mutate(across(starts_with("day"), ~ as.numeric(.)))

total_votes_by_gender <- vote_early_data |> 
  group_by(gender_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop')

total_votes_by_gender
```

**Age group**

```{r}
vote_early_data
```

```{r}
total_votes_age <- vote_early_data |> 
  group_by(age_group) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes))

total_votes_age
```

```{r}
ggplot(total_votes_age, aes(x = reorder(age_group, -total_votes), y = total_votes)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Total Early Votes by Age Group",
       x = "Age Group",
       y = "Total Votes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**Party**

```{r}
party_votes <- vote_early_data |> 
  group_by(party_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes))

party_votes
```

```{r}
ggplot(party_votes, aes(x = reorder(party_code, -total_votes), y = total_votes)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(title = "Total Early Votes by Party",
       x = "Party Code",
       y = "Total Votes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
gender_party_2024 <- vote_early_data |> 
  group_by(gender_code, party_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes))

gender_party_2024

write.csv(gender_party_2024, "gender_party_2024.csv", row.names = FALSE)

```








**comparing early votes for 2024 and early votes for 2016**

```{r}
early_votes_2016 <- read_csv("PG16 EarlyVoting RAW data.csv")
early_votes_2016
```
```{r}
early_votes_2016 <- early_votes_2016 |>
  clean_names()

early_votes_2016
```
 
votes for each county for each day in 2016
```{r}
long_data <- early_votes_2016 |> 
   pivot_longer(cols = starts_with("day"), 
              names_to = "day", 
              values_to = "votes")

# Calculate the total votes per county for each day
early_votes_county_day_2016 <- long_data |> 
   group_by(county_name, day) |> 
   summarize(total_votes = sum(votes, na.rm = TRUE), .groups = 'drop')

# View the results
print(early_votes_county_day_2016)
```
**county votes 2016**
```{r}
county_totals_2016<- early_votes_county_day_2016 |> 
  group_by(county_name) |> 
  summarise(total_early_votes = sum(total_votes)) |> 
  arrange(desc(total_early_votes))
  
county_totals_2016
```

```{r}
census_data_2016 <- get_acs(
  geography = "county",
  variables = "B01001_026",
  year = 2016,
  state = "MD",
  survey = "acs5"
)

census_data_2016 <- census_data_2016 |> 
  rename(county_name = NAME, population = estimate)
  

print(census_data_2016)
```

```{r}
census_data_2016 <- census_data_2016 |>
  mutate(county_name = if_else(
    county_name %in% c("Baltimore City", "Baltimore County"),
    county_name,
    str_replace(county_name, " County, Maryland", "") 
  ))

census_data_2016
```

```{r}
census_data_2016 <- census_data_2016 |> 
  mutate(county_name = case_when(
    county_name == "Baltimore" ~ "Baltimore County",
    county_name == "Baltimore city, Maryland" ~ "Baltimore City",
    county_name == "St. Mary's" ~ "Saint Mary's", 
    TRUE ~ county_name  
  ))

census_data_2016
```
```{r}
county_pop_2016_early_vote <- county_totals_2016 |> 
  left_join(census_data_2016, by = "county_name")

county_pop_2016_early_vote
```
```{r}
county_2016_pop_vote <- county_pop_2016_early_vote |> 
  mutate(votes_per_10000 = (total_early_votes / population) * 10000) |> 
  arrange(desc(votes_per_10000))

county_2016_pop_vote
```

```{r}
ggplot(county_2016_pop_vote, aes(x = reorder(county_name, votes_per_10000), y = votes_per_10000)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Early Votes per 10,000 People by County",
    x = "County",
    y = "Votes per 10,000 People"
  ) +
  theme_minimal()
```
**party/county early votes analysis 2016
```{r}
party_county_2016_votes <- early_votes_2016 |> 
   filter(party_code %in% c("DEM", "REP", "UNA")) |> 
  group_by(county_name, party_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes)) 


total_2016_votes_per_capita <- party_county_2016_votes |> 
  left_join(census_data_2016, by = "county_name") |> 
  mutate(votes_per_10000 = (total_votes / population) * 10000) |> 
  select(county_name, party_code, total_votes, votes_per_10000)

total_2016_votes_per_capita

```

```{r}
ggplot(total_2016_votes_per_capita, aes(x = county_name, y = votes_per_10000, fill = party_code)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Early Votes per 10,000 Residents by County and Party in 2016",
       x = "County",
       y = "Votes per 10,000 Residents") +
  scale_fill_manual(values = c("DEM" = "blue", "REP" = "red", "UNA" = "green")) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

**Age groups voting early in 2016**
```{r}
age_early_voters_2016 <- early_votes_2016 |> 
  group_by(age_group) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes))

age_early_voters_2016
```
```{r}
ggplot(age_early_voters_2016, aes(x = reorder(age_group, -total_votes), y = total_votes)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Total Early Votes by Age Group in 2016",
       x = "Age Group",
       y = "Total Votes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Gender demographics voting early in 2016**

```{r}
gender_early_voters_2016 <- early_votes_2016 |> 
  group_by(gender_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop')

gender_early_voters_2016
```

**Parties voting early in 2016**
```{r}
party_votes_2016 <- early_votes_2016 |>
  group_by(party_code) |> 
  summarise(total_votes = sum(day1, day2, day3, day4, day5, day6, day7, day8, na.rm = TRUE), .groups = 'drop') |> 
  arrange(desc(total_votes))

party_votes_2016
```
```{r}
ggplot(party_votes_2016, aes(x = reorder(party_code, -total_votes), y = total_votes)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(title = "Total Early Votes by Party in 2016",
       x = "Party Code",
       y = "Total Votes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Gender comparison between 2016 and 2024**
```{r}
merged_gender_votes <- total_votes_by_gender |> 
  left_join(gender_early_voters_2016, by = "gender_code")

merged_gender_votes <- merged_gender_votes |> 
  rename(
    gender_code_2024 = total_votes.x,
    gender_code_2016 = total_votes.y
  )

merged_gender_votes 
```

```{r}
total_votes_2024 <- sum(merged_gender_votes$gender_code_2024, na.rm = TRUE)
total_votes_2016 <- sum(merged_gender_votes$gender_code_2016, na.rm = TRUE)

merged_gender_votes <- merged_gender_votes |> 
  mutate(
    percent_2024 = (gender_code_2024 / total_votes_2024) * 100,
    percent_2016 = (gender_code_2016 / total_votes_2016) * 100
  )
merged_gender_votes
```

**Age comparison between 2016 and 2024**
```{r}
merged_age_votes <- total_votes_age |> 
  left_join(age_early_voters_2016, by = "age_group")

merged_age_votes <- merged_age_votes |> 
  rename(
    age_code_2024 = total_votes.x,
    age_code_2016 = total_votes.y
  )

merged_age_votes 
```

```{r}
total_votes_2024 <- sum(merged_age_votes$age_code_2024, na.rm = TRUE)
total_votes_2016 <- sum(merged_age_votes$age_code_2016, na.rm = TRUE)

merged_age_votes <- merged_age_votes |> 
  mutate(
    percent_2024 = (age_code_2024 / total_votes_2024) * 100,
    percent_2016 = (age_code_2016 / total_votes_2016) * 100
  )
merged_age_votes
```

**party comparison between 2016 and 2024**

```{r}
merged_party_votes <- party_votes |> 
  left_join(party_votes_2016, by = "party_code")

merged_party_votes <- merged_party_votes |> 
  rename(
    party_code_2024 = total_votes.x,
    party_code_2016 = total_votes.y
  )

merged_party_votes 
```

```{r}
total_votes_2024 <- sum(merged_party_votes$party_code_2024, na.rm = TRUE)
total_votes_2016 <- sum(merged_party_votes$party_code_2016, na.rm = TRUE)

merged_party_votes <- merged_party_votes |> 
  mutate(
    percent_2024 = (party_code_2024 / total_votes_2024) * 100,
    percent_2016 = (party_code_2016 / total_votes_2016) * 100
  )
merged_party_votes
```

**MEMO**

According to early voting data up to day 5, Oct. 28, women are leading in voting early at 315,869 with men at 251,338.

Democrats are leading in early voting followed by Republicans.

Age groups that are mostly voting early are 45-64 and 65 and older. 



Queen Anne's, Talbot, Calvert, Harford and Charles Counties have most early votes per 10,000 in the population. But Montgomery County is leading is most number of early votes. Democrats are leading in Montgomery County, PG County, Baltimore County, Baltimore City and Howard County.
